JavaScript as an Embedded DSL

### 2012-03-30 papers/ecoop2012/article.tex
global pass, intro and minor changes in different places

Developing rich web applications requires mastering different
environments on the client and server sides. While there is
considerable choice on the server-side, the client-side is tied to
JavaScript, which poses considerable software engineering challenges,
such as moving or sharing pieces of code between the environments. We
embed JavaScript as a DSL in Scala, using Lightweight Modular Staging.
DSL code can be compiled to JavaScript or executed as part of the
server application. We use features of the host language to make
client-side programming safer and more convenient. We use gradual
typing to interface typed DSL programs with existing JavaScript APIs.
We exploit a selective CPS transform already available in the host
language to provide a compelling abstraction over asynchronous
callback-driven programming in our DSL.

@@

One challenge in developing a large code base in JavaScript is the lack of static typing, as types are helpful for maintenance, refactoring, and reasoning about correctness. Furthermore, there is a need for more abstraction and modularity. ``Inversion of control'' in asynchronous callback-driven programming leads to code with control structures that are difficult to reason about. A challenge is to introduce helpful abstractions without a big hit on performance and/or code size. Communication between the server side and the client side aggravates the impedance mismatch: in particular, data validation logic needs to be duplicated on the client-side for interactivity and on the server-side for security.

@@

Another approach is to start with an existing language like Java, Scala or Clojure and compile it to JavaScript. Examples include GWT~\cite{gwt}, Scala+GWT~\cite{scala-gwt} and Clojurescript~\cite{clojurescript}. This approach addresses the problem of impedance mismatch between client and server programming but comes with its own set of challenges. In particular, compiling Scala code to JavaScript requires compiling Scala's standard library to JavaScript as any non-trivial Scala program uses Scala collections. This leads to not taking full advantage of libraries and abstractions provided by the target platform which results in big code size and suboptimal performance of Scala applications compiled to JavaScript. For example, a map keyed by strings would be implemented natively in JavaScript as an object literal, while, in Scala, one would likely use the hash map from the standard library, causing it to be compiled to and emulated in JavaScript. Moreover, both approaches tend to not accommodate very well to different API design and programming styles seen in many existing JavaScript libraries. The F\# Web Tools~\cite{tomasp-fsharp} is another example of starting with an existing language, F\#, and compiling it to both the client and the server sides. This work has similar goals to ours but the underlying techniques are very different (explicit quotations vs LMS, monads vs cps). A drawback of this second approach is that the whole starting language needs to be translated to JavaScript: there is no easy or modular way to limit the scope of the source language.

A third approach is to design a language than is a thin layer on top of JavaScript but provides some new features. A prime example of this idea is CoffeeScript~\cite{coffeescript}. This approach makes it easy to integrate with existing JavaScript libraries but does not solve the impedance mismatch problem. In addition, it typically does not give rise to new abstractions addressing problems seen in callback-driven programming style, though some JavaScript libraries such as Flapjax~\cite{flapjax} and Arrowlets~\cite{arrowlets} are specifically designed for this purpose.

@@

We propose to embed JavaScript as a DSL in a host language.\footnote{Surely, the embedded language is not exactly JavaScript: it naturally is a subset of Scala, the host language. However, it is quite close to JavaScript: often one can take snippets of JavaScript code and use them in the DSL with minor syntactic tweaking (as demonstrated by the Snowflake example described in section \ref{sec:trivial-embedding}).} Through LMS, we tackle the challenges outlined above with minimal effort, as most of the work is off-loaded to the host language. In particular, we make the following contributions:
\begin{itemize}
\item Our DSL is statically typed through the host language, yet supports gradual typing notably for incorporating external JavaScript libraries and APIs (section~\ref{sec:typing}).
\item In addition to generating JavaScript code, our DSL can be executed directly in the host language, allowing code to be shared between client and server (section~\ref{sec:trivial-embedding}).
\item We use advanced object-oriented techniques to achieve modularity in our DSL: each language primitive and API is defined in a separate module (section~\ref{sec:modularity}).
\item Our DSL supports typed object literals and class-based objects. The translations to JavaScript are lightweight and intuitive: the object literals translate to JSON-like object literals and the class-based objects to JavaScript constructor-based objects (section~\ref{sec:reification}).
\item With minimal effort, we exploit the selective CPS transform already existing in the host language to provide a compelling abstraction over asynchronous callback-driven programming in our DSL (section~\ref{sec:cps}). This case-study demonstrates the fruitfulness of re-using host language features to enhance our embedded DSL. A novel insight is that CPS transforming a program generator allows it to generate code that is in CPS.
\end{itemize}

@@

The present work extends both the range of virtualizable language features and the LMS framework, which is beneficial to future DSL efforts taking the JavaScript work as a case study. Previous LMS embeddings had to define each operation explicitly (like in section~\ref{sec:defineDslComponent}). This paper contributes lifting of whole traits or classes (through the \code{repProxy} mechanism described and used in sections~\ref{sec:typedApis}\&~\ref{sec:reified-classes}), untyped or optionally typed operations (sections~\ref{sec:casting},~\ref{sec:dynamic}\&~\ref{sec:dynamic2static}), and typed object literals (section~\ref{sec:object-literals}) including the necessary Scala-Virtualized support.

\section{Introduction to LMS}

@@

For \code{prog2}, notice that the \code{if} got transformed into an IR
node. To achieve this, LMS uses an extension of Scala,
Scala-Virtualized~\cite{scala-virtualized}, in which control
structures such as \code{if} can be reified into method calls, so that
alternative implementations can be provided. In our case, we provide an
implementation of \code{if} that constructs an IR node instead of
acting as a conditional. In addition, the \code{+} operation is
overloaded to act on both staged and unstaged expressions. This is
achieved by an implicit conversion from \code{Rep[Int]} to a class
\code{IntOps}, which defines a \code{+} method that creates an IR node
\code{Plus} when executed. Both of \code{Plus}'s arguments must be
staged. We use an implicit conversion to stage constants when needed
by creating a \code{Const} IR node.

@@

The generated code ressembles single-assignment form. The nested
\code{for}-loop is desugared into a \code{flatMap} which generates the
nested \code{for}-loop and the \code{splice} pattern concatenating the
inner \code{x4} arrays into one \code{x6} array in the JavaScript
code.\footnote{Obviously, the generated code can be optimized further.}

@@

In general, the framework provides an implicit conversion from
\code{Def} to \code{Exp}, which performs common subexpression
elimination by re-using the same symbol for identical definitions. We
don't use it here, because \code{log} is a side-effecting operation,
and we don't want to (re)move any such calls even if their message
is the same.

@@

We've implemented a component that generates JavaScript code that
asserts casts at runtime. It was fairly straightforward as the host
language allows us to easily inspect types involved in casting. Since
this component just provides a different implementation of the same
casting method \code{as}, it can be enabled selectively for
performance reasons.

@@

\code{DynamicRep} exploits a new Scala feature\footnote{C\# 4.0 type ``dynamic''~\cite{csharp-dynamic} can serve the same purpose.}, based on a
special trait \code{Dynamic}: An expression whose type \code{T} is a
subtype of \code{Dynamic} is subject to the following rewrites:
\begin{itemize}
\item \code{x.f} rewrites to \code{x.selectDynamic("f")},
\item \code{x.m(a, ..., z)} rewrites to \code{x.applyDynamic("m")(a, ..., z)},
\item \code{x.f = v} rewrites to \code{x.updateDynamic("f")(v)}.
\end{itemize}
These rewrites take place when the type \code{T} doesn't statically
have field \code{f} and method \code{m}.

@@

In the JavaScript example, when \code{abs} gets evaluated, through mixing in \code{JSExp}, it results in an IR tree roughly equivalent to {\tt\small If(LessThan(Sym("x"), Const(0)), Neg(Sym("x")), Sym("x"))}. In the trivial embedding, when \code{abs(-3)} gets called, it evaluates to \code{3} by executing the virtualized \code{if} as a normal condition. In short, in the trivial embedding, the virtualized method calls are evaluated in-place without constructing IR nodes.

@@

A drawback of the trivial embedding is that the virtualization overhead is incurred each time the program is evaluated. To avoid this, we could generate Scala code in the same way as we now generate JavaScript code, relying on a mechanism to incorporate the generated Scala code into the rest of the program. In fact, this approach is taken by the Delite framework.

@@

We also want the ability to treat functions as first-class staged
values, since JavaScript supports them. In order to do this, we
provide a higher-order function \code{fun} which takes a function of
type \code{Rep[A] => Rep[B]} and converts\footnote{We refer the reader
  to ~\cite{lms} for implementation details} it to a staged function
of type \code{Rep[A => B]}. For example, if we define \code{inc} in
the following way, its type is \code{Rep[Int => Int]}:
\begin{lstlisting}
val inc = fun { (x: Rep[Int]) => x + 1 }
\end{lstlisting}
Calling \code{inc(4*x)} results in an \code{Apply} IR node. We
actually generate JavaScript code for the staged \code{inc} function,
while we didn't for the unstaged one, since it's inlined at every call
site during staging. The generated code looks roughly like the
following:
\begin{lstlisting}[language=JavaScript]
var inc = function(a) {
  return a+1
}
inc(4*x)
\end{lstlisting}

@@

As an alternative to the callback-driven programming style, one can use an explicit monadic style, possibly sugared by a Haskell-like ``do''-notation. However, this requires rewriting possibly large parts of a program into monadic style when a single async operation is added. Another possibility is to automatically transform the program into continuation passing style (CPS), enabling the programmer to express the algorithm in a straightforward, sequential way and creating all the necessary callbacks and book-keeping code automatically. Links~\cite{links} uses this approach. However, a whole-program CPS transformation can cause performance degradation, code size blow-up, and stack overflows. In addition, it is not clear how to interact with existing non-CPS libraries as the whole program needs to adhere to the CPS style. We suggest using a {\it selective} CPS transformation, which precisely identifies what needs to be CPS transformed.

In fact, the Scala compiler already does selective, \code{@suspendable} type-annotation-driven CPS transformations of Scala programs~\cite{cps-scala,danvy90,danvy92}. We show how this mechanism can be used for transforming our DSL code before staging and stripping out most CPS abstractions at staging time. The generated JavaScript code doesn't have any CPS-specific code but is written in CPS-style by use of JavaScript anonymous functions.

Our implementation to support continuations is an example of an interesting technique of applying selective cps
transformation to embedded DSLs by means of a deep linguistic reuse, exploiting a host language feature to
implement the corresponding DSL feature without (much) additional work. An interesting insight is that CPS
transforming a program generator allows it to generate code that is in CPS.

@@

As an example, we'll consider a \code{sleep} method implemented in non-blocking, asynchronous style. This is useful, for example, when using \code{ThreadPool}s as no thread is being blocked during the sleep period. Let's see how our \code{sleep} method written in CPS can be used:

@@

It's important to note that reification preserves function composition (roughly speaking it's homomorphism between DSL code and generated JavaScript code).
Specifically, let \code{f: Rep[A] => Rep[B]} and \code{g: Rep[B] => Rep[C]} then {\tt\small fun(g compose f) == (fun(g) compose fun(f))}
where we consider two reified functions to be equal if they yield the same observable effects at runtime
\footnote{Note that composition on the left and the right hand side of the equation is not the same operation but they have the same observable runtime effect.}.
That property of function reification is at the core of reusing the continuation monad in our DSL. Thanks to the fact that the continuation monad composes functions, we can just insert reification at some places (like in a \code{sleep}) and make sure that we reify \emph{effects} of the continuation monad without the need to reify the monad itself.

@@

It's interesting to note that, since the request already has the right structure for the \code{jQuery.ajax} function, we can simply pass it to the framework-provided \code{quote} method, which knows how to generate JavaScript representations of any \code{JSLiteral}.

@@

In fact, if we change the code to use \code{suspendable} instead of \code{parSuspendable} and run the generated JavaScript program, we'll get following output printed to the JavaScript console:
\begin{lstlisting}
fetching gkossakowski
finished fetching gkossakowski
fetched [...]
fetched [...]
fetching odersky
finished fetching odersky
fetched [...]
fetched [...]
fetching adriaanm
finished fetching adriaanm
fetched [...]
fetched [...]
done
\end{lstlisting}

@@

It's worth noting that \code{RichCellOps} is not reified so it will be dropped at staging time and its method will get inlined whenever used. Also, it contains CPS-specific code that allows us to capture the continuation. The \code{fun} function reifies the captured continuation.

@@

The only difference between the parallel and serial versions of the Twitter example~\ref{code:twitter_example} is the use of \code{parSuspendable} instead of \code{suspendable} so the parallel implementation of the \code{foreach} method is used. The rest of the code stays the same. It's easy to switch between both versions, and users are free to make their choice according to their needs and performance requirements.

%Parallelism doesn't come for free as it involves more CPS rewritings and involves indirections introduced by our abstractions. However, since it's easy to switch between both versions, users are free to make those choices according to their needs and performance requirements.

@@

We've implemented our DSL in Scala.\footnote{The code can be found at
  \url{http://github.com/js-scala}} We used our DSL to
develop a few web applications, which are simple but not
trivial. First, we implemented a few drawing examples like the
snowflakes of figure~\ref{fig:snowflakes}. We extended the Twitter
example from section~\ref{sec:cps}, which presents the latest tweets
from selected users in an interactive way. In order to do so, we
incorporated a useful subset of the DOM API and jQuery library using our
typed APIs.

Also, we integrated our DSL with an existing web framework, Play 2.0.\footnote{\url{http://www.playframework.org}}
Our DSL can be used to define Play form validators that are executed on both the client and server sides.
We managed to achieve that without deep changes to how the framework works internally proving that our
DSL can be used as a library.

\begin{wrapfigure}{r}{5cm}
\includegraphics[scale=0.2]{drawing}
\end{wrapfigure}
Finally, we developed a collaborative drawing application, which
includes a server-side component (implemented using the Jetty web
server). We use web sockets to communicate between the server and
clients. Each client transmits drawing commands to the server, which
broadcasts them to all the clients. When a new client joins, the
server sends the complete drawing history to the new client, and the
client reconstructs the image by playing the commands. A very
simple improvement is to make the server execute the drawing
commands as well, and keep an up-to-date bitmap of the drawing -- this
can easily be achieved by using the trivial embedding described in
section~\ref{sec:trivial-embedding}. New clients then just obtain the
bitmap instead of replaying the history, which is potentially large.

@@

\bibliographystyle{abbrv}
\bibliography{refs,ppl}

>>>>

Developing rich web applications requires mastering different
environments on the client and server sides. While there is
considerable choice on the server-side, the client-side is tied to
JavaScript, which poses substantial software engineering challenges,
such as moving or sharing pieces of code between the environments. We
embed JavaScript as a DSL in Scala, using Lightweight Modular Staging.
DSL code can be compiled to JavaScript or executed as part of the
server application. We use features of the host language to make
client-side programming safer and more convenient. We use gradual
typing to interface typed DSL programs with existing JavaScript APIs.
We exploit a selective CPS transform already available in the host
language to provide a compelling abstraction over asynchronous
callback-driven programming in our DSL.

@@

One challenge in developing a large code base in JavaScript is the lack of static typing, as types are helpful for maintenance, refactoring, and reasoning about correctness. Furthermore, there is a need for more abstraction and modularity. ``Inversion of control'' in asynchronous callback-driven programming leads to code with control structures that are difficult to reason about. Another challenge is to introduce helpful abstractions without a big hit on performance and/or code size. Communication between the server side and the client side aggravates the impedance mismatch: in particular, data validation logic needs to be duplicated on the client-side for interactivity and on the server-side for security.

@@

Another approach is to start with an existing language like Java, Scala or Clojure and compile it to JavaScript. Examples include GWT~\cite{gwt}, Scala+GWT~\cite{scala-gwt} and Clojurescript~\cite{clojurescript}. This approach addresses the problem of impedance mismatch between client and server programming but comes with its own set of challenges. In particular, compiling Scala code to JavaScript requires compiling Scala's standard library to JavaScript as any non-trivial Scala program uses Scala collections. This leads to not taking full advantage of libraries and abstractions provided by the target platform which results in big code size and suboptimal performance of Scala applications compiled to JavaScript. For example, a map keyed by strings would be implemented natively in JavaScript as an object literal, while, in Scala, one would likely use the hash map from the standard library, causing it to be compiled to and emulated in JavaScript. Moreover, both approaches tend to not accommodate very well to different API design and programming styles seen in many existing JavaScript libraries. 
%The F\# Web Tools~\cite{tomasp-fsharp} is another example of starting with an existing language, F\#, and compiling it to both the client and the server sides. This work has similar goals to ours but the underlying techniques are very different (explicit quotations vs LMS, monads vs cps). 
A drawback of this second approach is that the whole starting language needs to be translated to JavaScript: there is no easy or modular way to limit the scope of the source language.
The F\# Web Tools~\cite{tomasp-fsharp} are an interesting variation of this approach that employ F\# quotations to translate only
parts of a program and allow programmers to define custom mappings for individual data types.

A third approach is to design a language that is a thin layer on top of JavaScript but provides some new features. A prime example of this idea is CoffeeScript~\cite{coffeescript}. This approach makes it easy to integrate with existing JavaScript libraries but does not solve the impedance mismatch problem. In addition, it typically does not give rise to new abstractions addressing problems seen in callback-driven programming style, though some JavaScript libraries such as Flapjax~\cite{flapjax} and Arrowlets~\cite{arrowlets} are specifically designed for this purpose.

@@

We propose to embed JavaScript as a DSL in a host language. \footnote{Surely, the embedded language is not \emph{exactly} JavaScript: it naturally is a subset of Scala, the host language. However, it is quite close to JavaScript. Often one can take snippets of JavaScript code and use them in the DSL with minor syntactic tweaking, as demonstrated by the Snowflake example described in section \ref{sec:trivial-embedding}.}
Through LMS (reviewed in section~\ref{sec:lms}), we tackle the challenges outlined above with minimal effort, as most of the work is off-loaded to the host language. In particular, we make the following contributions:
\begin{itemize}
\item Our DSL is statically typed through the host language, yet supports gradual typing notably for incorporating external JavaScript libraries and APIs (section~\ref{sec:typing}).
\item In addition to generating JavaScript code, our DSL can be executed directly in the host language, allowing code to be shared between client and server (section~\ref{sec:trivial-embedding}).
\item We use advanced object-oriented techniques to achieve modularity in our DSL: each language primitive and API is defined in a separate module (section~\ref{sec:modularity}).
\item Our DSL supports typed object literals and class-based objects. The translations to JavaScript are lightweight and intuitive: the object literals translate to JSON-like object literals and the class-based objects to JavaScript constructor-based objects (section~\ref{sec:reification}).
\item On top of the straightforward embedding, we implement advanced abstractions in the host language.
With minimal effort, we exploit the selective CPS transform already existing in Scala to provide a compelling abstraction over asynchronous callback-driven programming in our DSL (section~\ref{sec:cps}). The new insight here is that CPS transforming a program generator allows it to generate code that is in CPS. This case-study demonstrates the fruitfulness of re-using existing host language features to enhance our embedded DSL. 
\end{itemize}

@@

In addition to the contributions above, the present work significantly extends the LMS framework, which is beneficial to future DSL efforts taking the JavaScript work as a case study. Previous LMS embeddings had to define each staged operation explicitly (like in section~\ref{sec:defineDslComponent}). This paper contributes lifting of whole traits or classes (through the \code{repProxy} mechanism described and used in sections~\ref{sec:typedApis}\&~\ref{sec:reified-classes}), untyped or optionally typed operations (sections~\ref{sec:casting},~\ref{sec:dynamic}\&~\ref{sec:dynamic2static}), and typed object 
literals (section~\ref{sec:object-literals}) including necessary language support of the 
Scala-Virtualized~\cite{scala-virtualized} compiler.

\section{Introduction to LMS}
\label{sec:lms}

@@

For \code{prog2}, notice that the \code{if} got transformed into an IR
node. To achieve this, LMS uses
Scala-Virtualized~\cite{scala-virtualized}, a suite of minimal
extensions to the regular Scala compiler, in which control
structures such as \code{if} can be reified into method calls, so that
alternative implementations can be provided. In our case, we provide an
implementation of \code{if} that constructs an IR node instead of
acting as a conditional. In addition, the \code{+} operation is
overloaded to act on both staged and unstaged expressions. This is
achieved by an implicit conversion from \code{Rep[Int]} to a class
\code{IntOps}, which defines a \code{+} method that creates an IR node
\code{Plus} when executed. Both of \code{Plus}'s arguments must be
staged. We use an implicit conversion to stage constants when needed
by creating a \code{Const} IR node.

@@

The generated code resembles single-assignment form. The nested
\code{for}-loop is desugared into a \code{flatMap} which generates the
nested \code{for}-loop and the \code{splice} pattern concatenating the
inner \code{x4} arrays into one \code{x6} array in the JavaScript
code.\footnote{Obviously, the generated code can be optimized further.}

@@

In general, the framework provides an implicit conversion from
\code{Def} to \code{Exp}, which performs common subexpression
elimination by re-using the same symbol for identical definitions. We
do not use the automatic conversion here, because \code{log} is a 
side-effecting operation, and we do not want 
to (re)move any such calls even if their message
is the same.

@@

We have implemented a component that generates JavaScript code that
asserts casts at runtime. It was fairly straightforward as the host
language allows us to easily inspect types involved in casting. Since
this component just provides a different implementation of the same
casting method \code{as}, it can be enabled selectively for
performance reasons.

@@

\code{DynamicRep} exploits a new Scala feature\footnote{C\#'s type ``dynamic''~\cite{csharp-dynamic} can serve the same purpose.}, 
based on a special trait \code{Dynamic}: An expression whose type \code{T} is a
subtype of \code{Dynamic} is subject to the following rewrites:
\begin{itemize}
\item \code{x.f} rewrites to \code{x.selectDynamic("f")},
\item \code{x.m(a, ..., z)} rewrites to \code{x.applyDynamic("m")(a, ..., z)},
\item \code{x.f = v} rewrites to \code{x.updateDynamic("f")(v)}.
\end{itemize}
These rewrites take place when the type \code{T} doesn't statically
have field \code{f} and method \code{m}.

@@

In the JavaScript example (when mixing in \code{JSExp}) evaluating \code{abs(x)} results in an IR tree roughly equivalent to {\tt\small If(LessThan(Sym("x"), Const(0)), Neg(Sym("x")), Sym("x"))}. In the trivial embedding, when \code{abs(-3)} gets called, it evaluates to \code{3} by executing the virtualized \code{if} as a normal condition. In short, in the trivial embedding, the virtualized method calls are evaluated in-place without constructing IR nodes.

@@

A drawback of the trivial embedding is that although lightweight, virtualization is not completely free, and a small virtualization overhead is incurred each time the program is evaluated. To avoid this, we could generate Scala code in the same way as we now generate JavaScript code, relying on a mechanism to incorporate the generated Scala code into the rest of the program. In fact, this approach is taken by the Delite framework.

@@

We also want the ability to treat functions as first-class staged
values, since JavaScript supports them. In order to do this, we
provide a higher-order function \code{fun} which takes a function of
type \code{Rep[A] => Rep[B]} and converts\footnote{We refer the reader
  to~\cite{lms} for implementation details.} it to a staged function
of type \code{Rep[A => B]}. For example, if we define \code{inc} in
the following way, its type is \code{Rep[Int => Int]}:
\begin{lstlisting}
val inc = fun { (x: Rep[Int]) => x + 1 }
\end{lstlisting}
Calling \code{inc(4*x)} results in an \code{Apply} IR node. We
actually generate JavaScript code for the staged \code{inc} function,
while we did not for the unstaged one, since it is inlined at every call
site during staging. The generated code looks roughly like the
following:
\begin{lstlisting}[language=JavaScript]
var inc = function(a) {
  return a+1
}
inc(4*x)
\end{lstlisting}

@@

As an alternative to the callback-driven programming style, one can use an explicit monadic style, possibly sugared by a Haskell-like ``do''-notation. This is the approach taken by F\# Web Tools~\cite{tomasp-fsharp}. However, this requires rewriting possibly large parts of a program into monadic style when a single async operation is added. Another possibility is to automatically transform the program into continuation passing style (CPS), enabling the programmer to express the algorithm in a straightforward, sequential way and creating all the necessary callbacks and book-keeping code automatically. Links~\cite{links} uses this approach. However, a whole-program CPS transformation can cause performance degradation, code size blow-up, and stack overflows. In addition, it is not clear how to interact with existing non-CPS libraries as the whole program needs to adhere to the CPS style. We suggest using a {\it selective} CPS transformation, which precisely identifies what needs to be CPS transformed.

In fact, the Scala compiler already does selective CPS transformations of Scala programs, driven by \code{@suspendable} type annotations ~\cite{cps-scala,danvy90,danvy92}. We show how this mechanism can be used for transforming our DSL code before staging and stripping out most CPS abstractions at staging time. The generated JavaScript code does not contain any CPS-specific code but is written in CPS-style by use of JavaScript anonymous functions.

Our implementation to support continuations is an example of an interesting technique of applying selective cps
transformation to embedded DSLs by means of a deep linguistic reuse, exploiting a host language feature to
implement the corresponding DSL feature without (much) additional work. An interesting insight is that our method of
CPS transforming a direct-style program generator allows it to generate code that is in CPS.

@@

As an example, we consider a \code{sleep} method implemented in non-blocking, asynchronous style. This is useful, for example, when using \code{ThreadPool}s as no thread is being blocked during the sleep period. Let's see how our \code{sleep} method written in CPS can be used:

@@

It is important to note that reification preserves function composition (roughly speaking it is a homomorphism between DSL code and generated JavaScript code).
Specifically, let \code{f: Rep[A] => Rep[B]} and \code{g: Rep[B] => Rep[C]} then {\tt\small fun(g compose f) == (fun(g) compose fun(f))}
where we consider two reified functions to be equal if they yield the same observable effects at runtime
\footnote{Note that composition on the left and the right hand side of the equation is not the same operation but they have the same observable runtime effect.}.
That property of function reification is at the core of reusing the continuation monad in our DSL. Thanks to the fact that the continuation monad composes functions, we can just insert reification at some places (like in a \code{sleep}) and make sure that we reify \emph{effects} of the continuation monad without the need to reify the monad itself.

@@

It is interesting to note that, since the request already has the right structure for the \code{jQuery.ajax} function, we can simply pass it to the framework-provided \code{quote} method, which knows how to generate JavaScript representations of any \code{JSLiteral}.

@@

In fact, if we change the code to use \code{suspendable} instead of \code{parSuspendable} and run the generated JavaScript program, we will get following output printed to the JavaScript console:
\begin{lstlisting}
fetching gkossakowski
finished fetching gkossakowski
fetched [...]
fetched [...]
fetching odersky
finished fetching odersky
fetched [...]
fetched [...]
fetching adriaanm
finished fetching adriaanm
fetched [...]
fetched [...]
done
\end{lstlisting}

@@

It is worth noting that \code{RichCellOps} is not reified so it will be dropped at staging time and its method will get inlined whenever used. Also, it contains CPS-specific code that allows us to capture the continuation. The \code{fun} function reifies the captured continuation.

@@

The only difference between the parallel and serial versions of the Twitter example~\ref{code:twitter_example} is the use of \code{parSuspendable} instead of \code{suspendable} so the parallel implementation of the \code{foreach} method is used. The rest of the code stays the same. It is easy to switch between both versions, and users are free to make their choice according to their needs and performance requirements.

%Parallelism doesn't come for free as it involves more CPS rewritings and involves indirections introduced by our abstractions. However, since it is easy to switch between both versions, users are free to make those choices according to their needs and performance requirements.

@@

We have implemented our DSL in Scala.\footnote{The code can be found at
  \url{http://github.com/js-scala}} We used our DSL to
develop a few web applications, which are simple but not
trivial. First, we implemented a few drawing examples like the
snowflakes of figure~\ref{fig:snowflakes}. We extended the Twitter
example from section~\ref{sec:cps}, which presents the latest tweets
from selected users in an interactive way. In order to do so, we
incorporated a useful subset of the DOM API and jQuery library using our
typed APIs.

Moreover, we integrated our DSL with an existing web framework, Play 2.0.\footnote{\url{http://www.playframework.org}}
Our DSL can be used to define Play form validators that are executed on both the client and server sides.
We managed to achieve that without deep changes to the framework, proving that our
DSL can be used as a library.

\begin{wrapfigure}{r}{5cm}
\includegraphics[scale=0.2]{drawing}
\end{wrapfigure}
Finally, we developed a collaborative drawing application, which
includes a server-side component (implemented using the Jetty web
server). We use web sockets to communicate between the server and
clients. Each client transmits drawing commands to the server, which
broadcasts them to all the clients. When a new client joins, the
server sends the complete drawing history to the new client, and the
client reconstructs the image by playing back the commands. A very
simple improvement is to make the server execute the drawing
commands as well, and keep an up-to-date bitmap of the drawing -- this
can easily be achieved by using the trivial embedding described in
section~\ref{sec:trivial-embedding}. New clients then just obtain the
bitmap instead of replaying the history, which can be large and grow
unboundedly.

@@

{
%\clearpage
%\linespread{0.99}
\bibliographystyle{abbrv}
\bibliography{refs,ppl}
}
